---
title: "Spotify Data API Request"
author: "Rui Huang"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE, echo = TRUE, eval = FALSE)
```

# Abstract

This file is dedicated to obtain data through Spotify API requests. We will combine the httr package and the spotifyr package to get data, and the tidyverse package to tidy up data.


# API Requests

These requests are made through a combination of the httr package and spotifyr package.

```{r packages}
library(httr)
library(spotifyr)
library(tidyverse)
library(usethis)
```

#### Set Up API Key
```{r get api key}
clientID <- Sys.getenv('Spotify_clientID')
secret <- Sys.getenv('Spotify_Secret')

response <- POST(
  'https://accounts.spotify.com/api/token',
  accept_json(),
  authenticate(clientID, secret),
  body = list(grant_type = 'client_credentials'),
  encode = 'form',
  verbose()
)

mytoken <- content(response)$access_token

HeaderValue <- paste0('Bearer ', mytoken)

```


# Spotify Genres

Goal: Get a list of Spotify genres.

Reference: https://developer.spotify.com/console/get-available-genre-seeds/

```{r get genres}
URI <- paste0('https://api.spotify.com/v1/recommendations/available-genre-seeds')
response2 <- GET(url = URI, add_headers(Authorization = HeaderValue))
Genre <- content(response2)

genres <- map_chr(Genre$genres, ~.[[1]])

head(genres, 10)
```

# Artists by Genre

Goal: Request 50 artists per each genre we generated above through `get_genre_artists()` package.

```{r get artist by genre, results="hide"}
# Prepare an empty list to store artist data
artist_by_genre <- list()

# Loop through each genre and return 50 artist
for (i in seq_along(genres)) {
  artist_by_genre[[i]] <- get_genre_artists(genres[[i]], authorization = mytoken, limit = 50)
}
```

```{r tidy artists}
# Combine each artists data frames per genre
artist_df <- bind_rows(artist_by_genre)

# Tidy artist data
artist_df <- artist_df %>% 
  select(-genres, -href, -images, -type, -external_urls.spotify, -followers.href, -uri) %>%
  group_by(id) %>%
  mutate(genre_num = row_number(genre)) %>%
  ungroup() %>%
  pivot_wider(names_from = genre_num, names_prefix = 'genre_', values_from = genre) %>%
  rename(artist_id = id, artist_name = name) %>%
  rename(artist_id = aritst_id, artist_popularity = popularity, total_followers = followers.total, artist_genre = genre_1) %>%
  select(-genre_2:-genre_7)
 
# Take a look
head(artist_df)


# Save artists data to data folder
#write_csv(artist_df, '../data/artists.csv')
```

#### Get Artist Top Tracks

Goal: Request top tracks by each artist. On average, we will get approximately 10 
most famous tracks from an artist.

```{r top tracks by artist}
# Read artists data
artists <- read_csv('../data/artists.csv')

# Create an empty list to store track dataframe
top_track_list <- list()

# Store artist ids in a vector
artist_ids <- unique(artists$aritst_id)

# Request Top Tracks
tryCatch(
  {
    for (i in seq_along(artist_ids)) {
      top_track_list[[i]] <- get_artist_top_tracks(id = artist_ids[[i]], market = 'US', authorization = mytoken)
      Sys.sleep(0.1)
    }
  }, 
  error = function(error_message) {
    print(new_zero_popularity[i,]$track_id)
    print(i)
    message(error_message)
  }
)

# Convert the list to a single dataframe
top_tracks_by_artist <- bind_rows(top_track_list)

# Unnest artist ids and tidy data
top_tracks_c <- top_tracks_by_artist %>%
  select(-available_markets, -disc_number, -href, -is_local, -is_playable, -preview_url, -track_number, -type, -uri, -album.images, -album.type, -album.uri, -album.external_urls.spotify, -external_ids.isrc, -external_urls.spotify, -restrictions.reason, -album.href, -album.album_type, -album.artists, -album.total_tracks, -explicit) %>%
  rename(track_name = name, track_id = id, track_popularity = popularity) %>%
  unnest(artists) %>%
  select(-href, -type, -uri, -external_urls.spotify) %>%
  rename(artist_id = id, artist_name = name) %>%
  group_by(track_id) %>%
  mutate(artist_row_num = row_number()) %>%
  select(-artist_name) %>%
  pivot_wider(names_from = artist_row_num, names_prefix = 'artist_', values_from = artist_id) %>%
  ungroup(track_id) %>%
  inner_join(artists, by = c('artist_1' = 'aritst_id')) %>%
  select(track_id, track_name, track_popularity, artist_id = artist_1, album_id = album.id, album_name = album.name, album_release_date = album.release_date, album_release_date_precision = album.release_date_precision) %>%
  mutate(release_year = str_extract(album_release_date, '^\\d{4}')) %>%
  select(track_id, track_name, track_popularity, artist_id, release_year, album_id, album_name)

top_tracks_c

# write_csv(top_tracks_c, '../Data/top_tracks.csv')
```

#### Get Audio Features for Top Tracks

Goal: Require audio features for each track we obtained above.

```{r audio features for top tracks}
# Request Song popularity
tryCatch(
  {
    for (i in 0:373) {
        if (i == 0) {
          top_track_audio_features <- get_track_audio_features(ids = top_tracks$track_id[1:100], authorization = mytoken)
        } else if (1 <= i && i < 373) {
          top_track_audio_features <- bind_rows(top_track_audio_features, get_track_audio_features(ids = top_tracks$track_id[(i*100+1):(i*100+1+99)], authorization = mytoken))
          Sys.sleep(0.1)
        } else {
          top_track_audio_features <- bind_rows(top_track_audio_features, get_track_audio_features(ids = top_tracks$track_id[(i*100+1):nrow(top_tracks)], authorization = mytoken))
        }
      }
    }, 
  error = function(error_message) {
    print(i)
    message(error_message)
  }
)

# Tidy top_track_audio_features
top_track_audio_features <- top_track_audio_features %>%
  select(-type, -uri, -track_href, -analysis_url) %>%
  select(track_id = id, danceability:tempo, duration_ms, time_signature)

#write_csv(top_track_audio_features, '../data/top_track_audio_features.csv')

```


# Prepare Final Data Sets

Goal: Combine data sets to prepare final analysis data. We will have 2 data sets
at the end: tracks.csv and artists.csv.

```{r final data prep}
# Read datasets
artists <- read_csv('../data/artists.csv')
top_track_audio_features <- read_csv('../data/top_track_audio_features.csv')
top_tracks <- read_csv('../data/top_tracks.csv')

# Prepare artist data set
artists_c <- top_tracks %>%
  distinct(artist_id) %>%
  inner_join(artists, by = 'artist_id')
#write_csv(artists_c, '../output/artists.csv')

# Prepare track data set
tracks_c <- top_tracks %>%
  inner_join(top_track_audio_features, by = 'track_id') %>%
  select(track_id:track_popularity, release_year, danceability:time_signature, artist_id)
#write_csv(tracks_c, '../output/tracks.csv')

```
















